A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-01 tidyindex.qmd
rearrange and organise after meeting

## How does the pipeline constructin of an index look like? {#sec-toy-example}

Consider a commonly used drought index: Standardized Precipitation-Evapotranspiration Index (SPEI)  [@spei]. Its construction involves:

  1) transform the average temperature (`TMED`) into potential evapotranspiration (`pet`) 
  2) combine precipitation (`prcp`) and potential evapotranspiration (`pet`) into a single variable `diff`
  3) aggregate the difference series with a sliding window (`.scale`) 
  4) fit a distribution to the aggregated series, and 
  5) derive the index value from the normal density values.

Conventionally approach may combine all these steps into a single function, with some level of modularity. However, these modules may only work for the selected index offered by the package.

Under the pipeline approach, analysts first need to identify which module each step belongs to.

Below shows the pseudocode for constructing SPEI with the pipeline:

````
DATA %>% 
  var_trans(.method = thornthwaite, Tave = TMED, ..., .new_name = "pet") %>%
  dim_red(diff = prcp - pet) %>%
  aggregate(.var = diff, .scale = 12, .new_name = "agg") %>%
  dist_fit(.method = "lmoms", .var = agg, .dist = DIST) %>%
  augment(.var = agg)
````

<!-- *FWI:*  -->

<!--   - *verbs as name for each step: `aggregate`, `normalise`* -->
<!--   - *use NSE for variable name: `prcp`, `c(tavg, prcp)`* -->
<!--   - *created variables are prefixed with dot: `.agg_prcp`* -->
<!--   - *avoid string and use function to avoid "gamma" vs. "gam": `gamma()`, `lmom()`* -->

The pipeline construct allows for multiple `.scales` and `.dist` to be evaluated in `aggregate()` and `dist_fit()` to compare index under different parameterisations. The result can then be passed into the `ggplot2` to crease visualisation. @fig-toy-example compares the SPEI calculated with two distributions (log-logistic and pearson III).

Apart from evaluating multiples parameters, the pipeline approach allows the the steps written for one index can be directly extrapolate to another index building within the pipeline. The flexibility of the pipeline also integrate well with other existing packages, for examples, fitting distributions using L-moment is commonly used when constructing drought indexes. The package `lmomco` provides general L-moment fits to a wide range of distributions and users can easily access to all the distributions within the pipeline.

<!-- additional benefit: -->

<!--   - Multiple scales and multiple distributions can be fitted using the `c(...)` syntax to compare index values constructed from different parameterisations;  -->
<!--   - Intermediate results can be checked after each step; and -->
<!--   - Additional steps and analysis can be wired into the workflow for index diagnostics and customised user need. -->

```{r out.width = "100%"}
knitr::include_graphics(here::here("figures/toy-example-spei.png"))
```

```{r fig-toy-example, eval = FALSE, echo = FALSE}
#| fig-align: center
#| fig-width: 10
#| fig-height: 3
#| fig-cap: "Standardised Precipitation Evapotranspiration Index (SPEI) calculated with two distribution fits in Step 3 described above: generalised logistic and pearson III distribution, using the `wichita` data from the package `SPEI`."
library(tidyindex)
library(lmomco)
library(ggplot2)
library(tsibble)
library(SPEI)
library(tidyverse)
library(lubridate)
data(wichita)

dt <- wichita %>%
  mutate(ym = make_yearmonth(YEAR, MONTH), id = 1) %>% 
  init(id = id, time = ym) %>%
  var_trans(.pet_method = "thornthwaite", .vars = TMED, .lat = 37.6475) %>%
  dim_red(diff = PRCP - pet) %>%
  aggregate(.var = diff, .scale = 12) %>%
  dist_fit(.dist = list(loglogistic(), pearsonIII()), .method = "lmoms", .var = .agg) %>%
  augment(.var = .agg)

dt$data %>%
  mutate(.dist = ifelse(.dist == "parglo", "Log-logistic", "PearsonIII")) %>% 
  ggplot(aes(x = ym, y = .index, color = .dist)) +
  geom_line() + 
   theme_bw() +
  theme(
    panel.grid = element_blank()
  ) +
  scale_x_yearmonth(name = "Year", date_break = "2 years", date_label = "%Y") +
  scale_color_brewer(palette = "Dark2", name = "distribution") + 
  ylab("SPEI")

@@

<!-- *constructing time series index should also be encapsulated in the framework* -->

*any index can be broken down into multiple steps and then we can do things with it: swap, change parameter, etc*

*variables == indicators*

@@

The construction of an index sometimes needs to consider information from neighbouring time periods.
The temporal processing is a general operator on the time dimension of the data in the form of

@@

Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of

@@

The purpose of variable transformation is to create variables that fits assumptions for further computing. These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline. Variable transformation is a general notion of a functional transformation on the variable:

@@

While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of

@@

\newpage

@@

Dimension reduction summarises the multivariate information into univariate, which can be denoted as:

@@

<!-- With a probability model imposed,  -->
Distribution fit can be seen as the model fitting in its simplest term. It can be represented by

@@

This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. 
Also, the use of a normal scale needs to be interpreted with caution. 
@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.

@@

Benchmarking sets a constant value to allow the constructed index to be compared across time.
Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.

@@

In public communication, the index values are usually accompanied by a categorical grade. The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes. 
The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:

@@

# Incorporating alternative methods into the pipeline components {#sec-dev}

>>>>

@@

@@

**Input: **

**Output: **

**Computation:** The construction of an index sometimes needs to consider information from neighbouring time periods.
The temporal processing is a general operator on the time dimension of the data in the form of

@@

**Input: **

**Output: **

**Computation:** Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of

@@

**Input: **

**Output: **

**Computation:** The purpose of variable transformation is to create variables that fits assumptions for further computing. These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline. Variable transformation is a general notion of a functional transformation on the variable:

@@

**Input: **

**Output: **

**Computation:** While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of

@@

@@

**Input: **

**Output: **

**Computation:** Dimension reduction summarises the multivariate information into univariate, which can be denoted as:

@@

**Input: **

**Output: **

**Computation:** Distribution fit can be seen as the model fitting in its simplest term. It can be represented by

@@

**Input: **

**Output: **

**Computation:** This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. 
Also, the use of a normal scale needs to be interpreted with caution. 
@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.

@@

**Input: **

**Output: **

**Computation:** Benchmarking sets a constant value to allow the constructed index to be compared across time.
Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.

@@

**Input: **

**Output: **

**Computation:** In public communication, the index values are usually accompanied by a categorical grade. The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes. 
The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:

@@

# Software design {#sec-dev}

### 2023-08-01 tidyindex.tex
rearrange and organise after meeting

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-toy-example}{%
\subsection{How does the pipeline constructin of an index look
like?}\label{sec-toy-example}}

Consider a commonly used drought index: Standardized
Precipitation-Evapotranspiration Index (SPEI) (Vicente-Serrano,
Beguería, and López-Moreno 2010). Its construction involves:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  transform the average temperature (\texttt{TMED}) into potential
  evapotranspiration (\texttt{pet})
\item
  combine precipitation (\texttt{prcp}) and potential evapotranspiration
  (\texttt{pet}) into a single variable \texttt{diff}
\item
  aggregate the difference series with a sliding window
  (\texttt{.scale})
\item
  fit a distribution to the aggregated series, and
\item
  derive the index value from the normal density values.
\end{enumerate}

Conventionally approach may combine all these steps into a single
function, with some level of modularity. However, these modules may only
work for the selected index offered by the package.

Under the pipeline approach, analysts first need to identify which
module each step belongs to.

Below shows the pseudocode for constructing SPEI with the pipeline:

\begin{verbatim}
DATA %>% 
  var_trans(.method = thornthwaite, Tave = TMED, ..., .new_name = "pet") %>%
  dim_red(diff = prcp - pet) %>%
  aggregate(.var = diff, .scale = 12, .new_name = "agg") %>%
  dist_fit(.method = "lmoms", .var = agg, .dist = DIST) %>%
  augment(.var = agg)
\end{verbatim}

The pipeline construct allows for multiple \texttt{.scales} and
\texttt{.dist} to be evaluated in \texttt{aggregate()} and
\texttt{dist\_fit()} to compare index under different parameterisations.
The result can then be passed into the \texttt{ggplot2} to crease
visualisation. \textbf{?@fig-toy-example} compares the SPEI calculated
with two distributions (log-logistic and pearson III).

Apart from evaluating multiples parameters, the pipeline approach allows
the the steps written for one index can be directly extrapolate to
another index building within the pipeline. The flexibility of the
pipeline also integrate well with other existing packages, for examples,
fitting distributions using L-moment is commonly used when constructing
drought indexes. The package \texttt{lmomco} provides general L-moment
fits to a wide range of distributions and users can easily access to all
the distributions within the pipeline.

\includegraphics[width=1\textwidth,height=\textheight]{figures/toy-example-spei.png}

@@

\emph{any index can be broken down into multiple steps and then we can
do things with it: swap, change parameter, etc}

\emph{variables == indicators}

@@

The construction of an index sometimes needs to consider information
from neighbouring time periods. The temporal processing is a general
operator on the time dimension of the data in the form of

@@

Spatial processing may be needed when indexes are not calculated
independently on each collected location or when variables collected
from multiple sources need to be fused before further processing. The
process can be written as a general operation in the form of

@@

The purpose of variable transformation is to create variables that fits
assumptions for further computing. These assumptions include a stable
variance, normal distribution, or a certain scale required by some
algorithms down the pipeline. Variable transformation is a general
notion of a functional transformation on the variable:

@@

While scaling can be seen as a specific type of variable transformation,
it is separated into its own step to make the step explicit in the
pipeline. The key difference between the two steps is that variable
transformation typically changes the shape of the data while scaling
only changes the data scale and can usually be written in the form of

@@

Dimension reduction summarises the multivariate information into
univariate, which can be denoted as:

@@

Distribution fit can be seen as the model fitting in its simplest term.
It can be represented by

@@

This step maps the univariate series into a different scale, typically
for ease of comparison across regions. For example, a normal scale,
{[}0, 1{]}, or {[}0, 100{]} may be favored for reporting certain
indexes. In drought indexes, i.e.~SPI or SPEI, the quantiles from the
fitted distribution are converted into the normal scale via the normal
reverse CDF function: \(\Phi^{-1}(.)\). Normalising is usually used at
the end of the pipeline and its main difference from the scaling step is
that here the change of scale also changes the distribution of the
variable. While being commonly used, this step can get criticism from
analysts for forcing the data into the decided scale, which can be
either unnecessary or inaccurately exaggerate or downplay the outliers.
Also, the use of a normal scale needs to be interpreted with caution.
Figure~\ref{fig-normalising} illustrates the normal density not being
directly proportional to its probability of occurrence. This is
concerning, especially at the extreme values, since a small difference
in the tail density can have magnitudes of difference in its probability
of occurrence.

@@

Benchmarking sets a constant value to allow the constructed index to be
compared across time. Here we denote it with \(u[x(s_i, t_j)]\) where
\(u\) is a scalar of interest in the index constructed. A benchmark
value could be a constant or a function of the data, i.e.~mean.

@@

In public communication, the index values are usually accompanied by a
categorical grade. The categorised grades are an ordered set of
descriptive words or colors to communicate the severity or guide the
comprehension of the indexes. The mapping from continuous index values
to the discrete grades is called simplification in the pipeline and it
can be written as a piece-wise function:

@@

\hypertarget{sec-dev}{%
\section{Incorporating alternative methods into the pipeline
components}\label{sec-dev}}

@@

The function \texttt{compute\_indexes()} can be used to collectively
compute multiple indexes. The tidyindex offers wrapper functions, with
the prefix idx\_, that simplify the calculation of commonly used indexes
by combining a set of pipeline steps into a single function. For
example, the function \texttt{idx\_spei()} includes the five steps
previously described in Section~\ref{sec-toy-example} (variable
transformation, dimension reduction, temporal aggregation, distribution
fit, and normalise). Each \texttt{idx\_xxx()} function specifies the
relevant parameters relevant to the index: the thornthwaite method is
used to calculate PET in SPEI, with the average temperature
(\texttt{tavg}) and latitude (\texttt{lat}) used as inputs. The SPEI is
computed at four time scales (6, 12, 24, and 36 months) and fitted with
two distributions (Log-logistic and General Extreme Value (GEV)). The
SPI is also computed at the same four time scales and uses the default
gamma distribution to fit the aggregated series.

@@

\leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
Vicente-Serrano, Sergio M., Santiago Beguería, and Juan I. López-Moreno.
2010. {``A {Multiscalar} {Drought} {Index} {Sensitive} to {Global}
{Warming}: {The} {Standardized} {Precipitation} {Evapotranspiration}
{Index}.''} \emph{Journal of Climate} 23 (7): 1696--1718.
\url{https://journals.ametsoc.org/view/journals/clim/23/7/2009jcli2909.1.xml}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

@@

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} The construction of an index sometimes needs to
consider information from neighbouring time periods. The temporal
processing is a general operator on the time dimension of the data in
the form of

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Spatial processing may be needed when indexes are
not calculated independently on each collected location or when
variables collected from multiple sources need to be fused before
further processing. The process can be written as a general operation in
the form of

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} The purpose of variable transformation is to
create variables that fits assumptions for further computing. These
assumptions include a stable variance, normal distribution, or a certain
scale required by some algorithms down the pipeline. Variable
transformation is a general notion of a functional transformation on the
variable:

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} While scaling can be seen as a specific type of
variable transformation, it is separated into its own step to make the
step explicit in the pipeline. The key difference between the two steps
is that variable transformation typically changes the shape of the data
while scaling only changes the data scale and can usually be written in
the form of

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Dimension reduction summarises the multivariate
information into univariate, which can be denoted as:

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Distribution fit can be seen as the model fitting
in its simplest term. It can be represented by

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} This step maps the univariate series into a
different scale, typically for ease of comparison across regions. For
example, a normal scale, {[}0, 1{]}, or {[}0, 100{]} may be favored for
reporting certain indexes. In drought indexes, i.e.~SPI or SPEI, the
quantiles from the fitted distribution are converted into the normal
scale via the normal reverse CDF function: \(\Phi^{-1}(.)\). Normalising
is usually used at the end of the pipeline and its main difference from
the scaling step is that here the change of scale also changes the
distribution of the variable. While being commonly used, this step can
get criticism from analysts for forcing the data into the decided scale,
which can be either unnecessary or inaccurately exaggerate or downplay
the outliers. Also, the use of a normal scale needs to be interpreted
with caution. Figure~\ref{fig-normalising} illustrates the normal
density not being directly proportional to its probability of
occurrence. This is concerning, especially at the extreme values, since
a small difference in the tail density can have magnitudes of difference
in its probability of occurrence.

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Benchmarking sets a constant value to allow the
constructed index to be compared across time. Here we denote it with
\(u[x(s_i, t_j)]\) where \(u\) is a scalar of interest in the index
constructed. A benchmark value could be a constant or a function of the
data, i.e.~mean.

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} In public communication, the index values are
usually accompanied by a categorical grade. The categorised grades are
an ordered set of descriptive words or colors to communicate the
severity or guide the comprehension of the indexes. The mapping from
continuous index values to the discrete grades is called simplification
in the pipeline and it can be written as a piece-wise function:

@@

\hypertarget{sec-dev}{%
\section{Software design}\label{sec-dev}}

@@

The function \texttt{compute\_indexes()} can be used to collectively
compute multiple indexes. The tidyindex offers wrapper functions, with
the prefix idx\_, that simplify the calculation of commonly used indexes
by combining a set of pipeline steps into a single function. For
example, the function \texttt{idx\_spei()} includes the five steps
previously described in \textbf{?@sec-toy-example} (variable
transformation, dimension reduction, temporal aggregation, distribution
fit, and normalise). Each \texttt{idx\_xxx()} function specifies the
relevant parameters relevant to the index: the thornthwaite method is
used to calculate PET in SPEI, with the average temperature
(\texttt{tavg}) and latitude (\texttt{lat}) used as inputs. The SPEI is
computed at four time scales (6, 12, 24, and 36 months) and fitted with
two distributions (Log-logistic and General Extreme Value (GEV)). The
SPI is also computed at the same four time scales and uses the default
gamma distribution to fit the aggregated series.

@@
